---
title: "Homework 2"
author: Emma Sexton <br>
date: "Due: 5 October 2022"
output: github_document
---

```{r, eval = FALSE}
library(tidyverse)
```

## Problem 1

First, we need to read and clean the data. 

We want to retain line, station name, station latitude / longitude, route served, entry, vending, entrance type, and ADA compliance. 

```{r}
#nyc_transit <- 
  #read_csv("data/NYC_transit_data.csv") %>% 
  #janitor::clean_names()
```

We also need to convert the `entry` variable from character to logical variable using `ifelse` or `recode`. 

```{r}

```


##### Summary of Dataset:


##### *Q1.1: How many distinct stations are there?*

##### *Q1.2: How many stations are ADA compliant?*

##### *Q1.3: What proportion of station entrances / exits without vending allow entrance?*


Next, we will reformat the data so that `route_number` and `route_name` are distinct variables. 

```{r}

```

##### *Q1.4: How many distinct stations serve the A train? Of the stations that serve the A train, how many are ADA compliant?*


