---
title: "Homework 2"
author: Emma Sexton <br>
date: "Due: 5 October 2022"
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, message = FALSE)
```

```{r}
library(tidyverse)
library(readxl)
```

## Problem 1

The goal of this problem is to import data and use the `select`, `filter`, `mutate`, and `pivot_longer` commands to summarize the dataset and answer data-specific questions. 

```{r}
nyc_transit_df <- 
  read_csv(
    'data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv',
    col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c")) %>%  
  janitor::clean_names() %>% 
  select(
    line, station_name, station_latitude, station_longitude, 
    starts_with("route"), entrance_type, entry, vending, ada) %>%  
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE))
```

##### Summary of Dataset:

Above, we imported the `NYC_Transit_Subway_Entrance_And_Exit_Data.csv` data which has 1868 rows (observations) and 20 columns (variables). We also converted `route` variables 8-11 to character types to be consistent with 1-7. Additionally, we retained the following variables: line, station name, station latitude / longitude, route served, entry, vending, entrance type, and ADA compliance. We also updated `entry` from being a character with `YES` and `NO` values to a logical vector. 

Although clean, the dataframe is not 'tidy'. Two new variables - `route_name` and `route_number` - need to be created to manipulate the `route` variables from wide to long format. This can be used to answer questions related to specific routes, but it may not be helpful for answering questions that focus on station-level variables as it'll repeat observations.

##### *Q1.1: How many distinct stations are there?*
To determine the number of distinct stations, we can select station name and line using the `distinct()` function. The number of rows in the dataset equals the number of unique stations. 

```{r}
unique_stations <- nyc_transit_df %>% 
  select(station_name, line) %>% 
  distinct
```

According to the chunk code above, there are **465 unique stations**. 

##### *Q1.2: How many stations are ADA compliant?*

To determine the number of stations that are ADA compliant, we can filter on the `ada` variable using `filter()`. 

```{r}
ada_compliant <- nyc_transit_df %>% 
  filter(ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct
```

According to the chunk code above, there are **84 stations that are ADA compliant**. 

##### *Q1.3: What proportion of station entrances / exits without vending allow entrance?*

```{r}
entrance_wo_vending <- nyc_transit_df %>% 
  filter(vending == "NO") %>% 
  pull(entry) %>% 
  mean
```

Here, we filtered on `vending == NO` and extracted a single column (here, `entry`). Taking the mean of a logical variable allows us to determine the proportion of entrances to exits, and we discover that **37.7% of station entrances/exits without vending allow entrance**. 

##### *Q1.4: How many distinct stations serve the A train? Of the stations that serve the A train, how many are ADA compliant?*

Next, we will reformat the data so that `route_number` and `route_name` are distinct variables. In tidying the data using `pivot_longer`, we are able to reformat the data from wide to long, which allows us to answer route specific questions as noted earlier. 

```{r}
transit_A_ada <- nyc_transit_df %>% 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name") %>% 
  filter(route_name == "A", ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct
```

In this chunk, we are pivoting the data and creating `route_number` and `route_name`, then filtering to choose the A train and ADA compliant stations. We then use the chunk code that we've used in previous questions to pull the distinct station names. There are **17 stations that serve the A train and are ADA compliant**. 


## Problem 2

The goal of this problem is to import, clean, and organize two datasets (`Mr. Trash Wheel` and `Professor Trash Wheel`), and then join the two datasets. 

We will start with importing, cleaning, and organizing `Mr. Trash Wheel`...

```{r}
mr_trashwheel <- 
  read_excel("data/Trash_Wheel_Collection_Data.xlsx", 
                            sheet = "Mr. Trash Wheel", 
                            range = "A2:N549") %>% 
  janitor::clean_names() %>% 
  mutate(sports_balls = round(sports_balls, digits = 0))
```


